### 2015-07-07 stochastic/chapter-3.Rmd
stochastic

During the time that ecologists have struggled to identify the interactions among dozens of species, bioinformaticians have largely solved the analogous problem of estimating interactions among *thousands* of genes from their co-expression levels (e.g. @friedman_sparse_2008). The discrepancy is due to the fact that ecologists have focused, almost exclusively, on overall co-occurrence rates [@connor_assembly_1979; @gotelli_empirical_2009; @gilpin_factors_1982; @veech_probabilistic_2013] or on closely-related values such as correlation coefficients [@pollock_understanding_2014; see @faisal_inferring_2010 for a recent exception]. As ecologists know, however, the correlation between two species will often reflect other factors beyond their direct pairwise interactions (Figure 1). When these factors are not accounted for, co-occurrence rates can reliably lead to incorrect inferences [@harris_estimating_2015]. While some methods have been proposed for incorporating a small number of specific factors (such as geographic or environmental dissimilarity) into ecologists' null models at a time [@lessard_strong_2011], we still don't have a good way to scale these approaches up for cases with many biotic and abiotic factors acting simultaneously.

@@

[[math for conditional distributions]]

These conditional probabilities can be combined to calculate the probability of observing any given combination of presences and absences [[Appendix?]].  For example, hypothetical assemblages including many pairs of competitively exclusive species at the same site will tend to be very improbable under the model.  Unfortunately, the extremely large number of possible assemblages makes this joint probability intractable when the number of species is large.  However, it is generally straightforward to simulate examples of assemblages that are consistent with graphical models such as these, using Monte Carlo techniques like Gibbs sampling [@salakhutdinov_efficient_2012]. Gibbs sampling is especially convenient for these models because it only requires the conditional probabilities, which are straightforward to compute (Harris 2015; Appendix).

@@

@@

To demonstrate that my stochastic approximation implementation could converge to the global optimum, I first simulated 10 landscapes with 20 species and 500 sites, using the Gibbs sampling approach described in @harris_estimating_2015.  These landscapes had few enough species that the `rosalia` package for estimating Markov networks [@alia_2015] could find the penalized maximum likelihood estimates for the Markov network in a reasonable amount of time.  For each of these simulated landscapes, I then fit the same model using both the exact approach from @harris_estimating_2015 and the stochastic approximation method described above.

@@

![](convergence.pdf)

@@

![](estimates.pdf)

@@

>>>>

During the time that ecologists have struggled to identify the interactions among dozens of species, bioinformaticians have largely solved the analogous problem of estimating interactions among *thousands* of genes from their co-expression levels (e.g. @friedman_sparse_2008). The discrepancy is due to the fact that ecologists have focused, almost exclusively, on overall co-occurrence rates [@connor_assembly_1979; @gotelli_empirical_2009; @gilpin_factors_1982; @veech_probabilistic_2013] or on closely-related values such as correlation coefficients [@pollock_understanding_2014; see @faisal_inferring_2010 for a recent exception]. As ecologists know, however, the correlation between two species will often reflect other factors beyond their direct pairwise interactions (Figure 1). When these factors are not accounted for, this reliance on overall co-occurrence rates can reliably lead to incorrect inferences [@harris_estimating_2015]. While some methods have been proposed for incorporating a small number of specific factors (such as geographic or environmental dissimilarity) into ecologists' null models at a time [@lessard_strong_2011], we still don't have a good way to scale these approaches up for cases with many biotic and abiotic factors acting simultaneously.

![A hypothetical network of three species, two of which depend on an abiotic factor (annual rainfall).  **A.** The “true” network involves the caterpillar exploiting the two tree species as well as competition between the two trees.  The two trees also benefit from rainfall.  **B.** Although the two trees directly reduce one another’s occurrence probabilities via competition, they tend to occur in the same areas (with sufficient water and without too many herbivores).  An analysis that ignores these other factors could mistakenly conclude that the two trees were mutualists because of their high co-occurrence rate.  **C.** The methods presented here are designed to estimate the direct interactions between each pair of species, after accounting for other species and abiotic factors that could affect their co-occurrence rates. Note that, with observational data, the bi-directional effects of species interactions (pairs of arrows between species) need to be collapsed to a single number describing the conditional relationship between species (one double-headed arrow).](figure-1.pdf)

@@

[[add math for conditional distributions; same as in chapter 2]]

These conditional probabilities can be combined to calculate the probability of observing any given combination of presences and absences [[Appendix?]].

[[Add math for joint distribution; same as in chapter 2]]

Unfortunately, the extremely large number of possible assemblages makes this joint probability intractable when the number of species is large.  However, it is generally straightforward to simulate examples of assemblages that are consistent with graphical models such as these, using Monte Carlo techniques like Gibbs sampling [@salakhutdinov_efficient_2012]. Gibbs sampling is especially convenient for these models because it only requires the conditional probabilities, which are straightforward to compute (Harris 2015; Appendix).

@@

## Conditioning the network on the abiotic environment

The Markov networks fit in @harris_estimating_2015 treated the $\alpha_i$ terms from Equations [[1 and 2]] as constants, meaning that the models expected each species to have the same conditional occurrence probability in any location with a given set of facilitators and competitors.  In this paper, I extend the model so that the local abiotic conditions can also affect species occurrence probabilities directly. Specifically, the $\alpha_i$ terms in these analyses are linear combinations of the local environmental conditions, represented by $x_1$ through $x_5$. By estimating the coefficients associated with each of these $x$ variables for each species, we can account for their responses to the abiotic environment as well as to one another.  In other fields, a model like this one, where a Markov network’s parameters depend on external factors is called a conditional random field [refs]; these models have been used in a variety of contexts, from language models to image processing [refs].

@@

To demonstrate that my stochastic approximation implementation could converge to the global optimum, I first simulated 10 landscapes with 20 species and 500 sites, using the Gibbs sampling approach described in @harris_estimating_2015.  These landscapes had few enough species that the `rosalia` package for estimating Markov networks [@rosalia_2015] could find the penalized maximum likelihood estimates for the Markov network in a reasonable amount of time.  For each of these simulated landscapes, I then fit the same model using both the exact approach from @harris_estimating_2015 and the stochastic approximation method described above.

@@

![Stochastic approximation decreases the model’s error over time. **A.** With small networks where the exact maximum likelihood estimates are known, stochastic approximation reaches the global optimum very quickly. **B.** With larger networks, where the “true” parameter values that generated the network are known but the maximum likelihood estimates for the observed landscape are not, stochastic approximation initially drives the error down very quickly, and then slowly approaches the apparent optimum.](convergence.pdf)

@@

![“True” network parameters for all 31,125 pairwise interactions versus the estimates from the three methods presented. Blue lines represent linear regressions through each cloud of points.](estimates.pdf)

@@

Richer forms of data, e.g. from observational time series or controlled experiments, could help ecologists recover more information about species interactions than is possible with the kinds of observational data analyzed here: in particular, these richer forms of data will be necessary to answer questions about which species in a positively associated pair is facilitating the other or which species in a negatively associated pair is excluding the other.
